---
title: "Selecting, Grouping, and Iterating Tiles"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Selecting, Grouping, and Iterating Tiles}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(tilework)
library(terra)

f <- system.file("ex/elev.tif", package = "terra")
r <- rast(f)

tp <- tilePlan("pixel")
tp$pxdims <- dim(r)[1:2]
tp$nrows   <- 20L
tp$ncols   <- 20L
```

## Overview

A `tilePlan` defines the geometry of a tiling. Three higher-level structures
control *which* tiles get processed and *how*:

| Class | Purpose | Parallelization |
|---|---|---|
| `tileSelection` | Lazy index subset of a plan | Same as underlying `tilePlan` |
| `tileGroup` | Named groups of tiles | Across groups **or** within groups |
| `tileIterator` | Stateful streaming in batches | Batches distributed across workers |

These are not mutually exclusive — a `tileGroup` can be wrapped in a
`tileIterator`, and a `tileGroup` itself wraps a `tilePlan`.

---

## Basic tile processing

Before introducing the higher-level structures, it helps to see what
`tileApply` looks like on a plain `tilePlan`. This is the baseline that
`tileSelection`, `tileGroup`, and `tileIterator` all build on.

```{r basic}
results <- tileApply(r, tiles = tp, FUN = function(x, .I, .R, .C) {
    list(
        mean = terra::global(x, "mean", na.rm = TRUE)[[1L]],
        tile = .I,
        row  = .R,
        col  = .C
    )
})

str(results[[1L]])
```

`FUN` receives the extracted tile data as its first argument. Four special
parameters are available and injected automatically when present in
`formals(FUN)`:

| Injectable | Type | Value |
|---|---|---|
| `.I` | `integer` | Flat tile index |
| `.R` | `integer` | Tile row in the grid |
| `.C` | `integer` | Tile column in the grid |
| `.TILE` | `SpatExtent` or `integer[4]` | Tile bounds object |

Every tile is processed independently. When a `future` plan with multiple
workers is active, tiles are distributed across workers automatically. The
higher-level structures below give you control over *which* tiles are
processed, in *what order*, and with *what per-worker state*.

---

## `tileSelection`

`tileSelection` is the result of `drop = FALSE` indexing on a `tilePlan`. It
stores a subset of tile indices without computing any bounds, and is passed
directly to `tileApply` like a plan.

```{r selection}
# Select specific tiles without materialising bounds
sel <- tp[c(1, 5, 9, 13), drop = FALSE]
sel
length(sel)
```

```{r selection-apply}
results <- tileApply(r, tiles = sel, FUN = function(x, .I) {
    terra::global(x, "mean", na.rm = TRUE)[[1L]]
})
str(results)
```

**When to use:** You have a plan but only want to process a known subset —
for example, tiles that intersect a region of interest, tiles flagged in
metadata, or a manual quality-control subset. It avoids rebuilding the plan
and keeps the original tile indices as identifiers.

---

## `tileGroup`

`tileGroup` organises tiles from a plan into named groups. Each group is a
logical partition — groups can overlap, and not all tiles need to be assigned
to a group.

```{r tilegroup}
nr <- nrow(tp)
nc <- ncol(tp)

tg <- tileGroup(tp, groups = list(
    interior = list(2:(nr - 1L), 2:(nc - 1L)),  # ij indexing
    border   = c(                                 # vector indexing
        # top and bottom rows
        seq_len(nc),
        seq(length(tp) - nc + 1L, length(tp)),
        # left and right columns (excluding corners)
        seq(nc + 1L, length(tp) - nc, by = nc),
        seq(nc + nc, length(tp) - 1L, by = nc)
    )
))

tg
```

Groups accept the same indexing styles as the underlying plan, and they can
be mixed within the same `tileGroup`:

- **Vector indexing** — a flat integer vector of tile indices.
- **ij indexing** — a `list(rows, cols)` that selects by grid position,
  expanding all row × col combinations. `list(2:(nr-1), 2:(nc-1))` means
  all interior rows × all interior cols.

### Parallelization strategies

`tileApply` on a `tileGroup` accepts a `parallel_strategy` argument:

**`"groups"` (default)** — groups are distributed across workers. Each worker
processes all tiles in its assigned group sequentially. This is the right
choice when:

- Groups have meaningfully different processing logic
- Per-worker setup cost is significant (model loading, DB connection) — use
  `setup_FUN` which runs once per group/worker
- Groups are large enough to keep workers busy

```{r par-groups, fig.show='hide'}
results_groups <- tileApply(r,
    tiles = tg,
    parallel_strategy = "groups",
    setup_FUN = function(.GROUP) {
        # Runs once per group — expensive setup goes here
        list(group_name = .GROUP, initialized_at = Sys.time())
    },
    FUN = function(x, .I, .GROUP, .SETUP_OUT) {
        list(
            tile  = .I,
            group = .GROUP,
            mean  = terra::global(x, "mean", na.rm = TRUE)[[1L]]
        )
    }
)

# Results are a named list with one element per group
names(results_groups)
length(results_groups$interior)
```

**`"tiles"`** — groups are processed sequentially, but tiles *within* each
group are distributed across workers. This is the right choice when:

- Each tile is expensive and independent
- Groups are few but tiles per group are many
- Per-worker setup cost is low (no `setup_FUN` available here)

```{r par-tiles}
results_tiles <- tileApply(r,
    tiles = tg,
    parallel_strategy = "tiles",
    FUN = function(x, .I, .GROUP) {
        terra::global(x, "mean", na.rm = TRUE)[[1L]]
    }
)

names(results_tiles)
```

### `group_FUN`

An optional `group_FUN` is applied to each group's result list after all tiles
in that group are processed. This is a convenient hook for within-group
aggregation, avoiding a separate post-processing pass.

```{r group-fun}
results_agg <- tileApply(r,
    tiles = tg,
    FUN = function(x, .I) {
        terra::global(x, "mean", na.rm = TRUE)[[1L]]
    },
    group_FUN = function(group_results, .GROUP) {
        vals <- unlist(group_results)
        list(
            group  = .GROUP,
            n      = length(vals),
            mean   = mean(vals, na.rm = TRUE),
            range  = range(vals, na.rm = TRUE)
        )
    }
)

str(results_agg, max.level = 2)
```

**When to use:** Any time tiles have a natural logical partition — different
tissue regions, different processing requirements, edge vs. interior tiles
(as in the [ML vignette](patch-feature-extraction.html)), or simply grouping
tiles for separate output files.

---

## `tileIterator`

`tileIterator` wraps any `tilePlan` or `tileGroup` (with `$active` set) and
adds stateful position tracking. Instead of processing all tiles at once, it
yields batches of a fixed size on demand.

```{r iterator}
iter <- tileIterator(tp, batch_size = 5L)
iter
```

### Manual streaming

For full control, iterate manually with `$has_next` and `$next_batch()`:

```{r manual-stream}
iter$reset()

while (iter$has_next) {
    batch <- getTile(r, iter)
    # process batch...
}

iter$progress  # 100%
```

### With `tileApply`

Passing a `tileIterator` to `tileApply` triggers the streaming dispatch path.
`iterSplit()` partitions the iterator's remaining range across
`future::nbrOfWorkers()` sub-iterators. Each worker independently streams
through its assigned tiles in batches.

```{r iter-apply}
iter$reset()

results <- tileApply(r,
    tiles = iter,
    FUN = function(batch, .I, .BATCH, .POSITION) {
        # batch is a list of SpatRaster tiles
        # .I    — tile indices in this batch
        # .BATCH — batch number within this worker
        # .POSITION — c(start_index, end_index) of this batch
        lapply(batch, function(tile) {
            terra::global(tile, "mean", na.rm = TRUE)[[1L]]
        })
    }
)
```

### `setup_FUN` for per-worker initialization

When `tileApply` distributes work across workers, `setup_FUN` runs once on
each worker before batch processing begins. Its output is passed to every
`FUN` call as `.SETUP_OUT`. This is the mechanism for loading a model,
opening a database connection, or any other expensive per-worker setup.

```{r setup-fun}
iter$reset()

results <- tileApply(r,
    tiles = iter,
    setup_FUN = function(.W, .X) {
        # .W — worker index
        # .X — the input data object
        list(worker = .W, nrow = nrow(.X), ncol = ncol(.X))
    },
    FUN = function(batch, .SETUP_OUT) {
        worker_info <- .SETUP_OUT
        lapply(batch, function(tile) {
            terra::global(tile, "mean", na.rm = TRUE)[[1L]]
        })
    }
)
```

### Iterator splitting

`iterSplit()` divides an iterator's remaining range into `n` independent
sub-iterators with no shared state. This is what `tileApply` calls internally,
but it is also available directly for custom parallel workflows.

```{r itersplit}
iter$reset()
sub_iters <- iterSplit(iter, n = 3L)

# Each sub-iterator covers a non-overlapping range
sapply(sub_iters, function(x) x$bound)
sapply(sub_iters, function(x) x$remaining)
```

**When to use:** Large datasets where loading all tile data simultaneously
would exhaust memory; workflows requiring checkpointing (serialize the
iterator mid-run and resume later); ML inference pipelines where batch size
must match the model's expected input count. See the
[ML feature extraction vignette](patch-feature-extraction.html) for a full
worked example.

---

## Choosing between them

| Situation | Recommended |
|---|---|
| Process a fixed known subset of tiles | `tileSelection` |
| Tiles have distinct logical groupings with different logic | `tileGroup` |
| Few groups, many tiles per group, tile processing is expensive | `tileGroup`, `parallel_strategy = "tiles"` |
| Many groups, tiles per group is small, setup cost per group matters | `tileGroup`, `parallel_strategy = "groups"` with `setup_FUN` |
| Dataset too large to hold all tile data in memory at once | `tileIterator` |
| Worker warmup is expensive (model loading) and batching is needed | `tileIterator` with `setup_FUN` |
| Checkpoint and resume a long-running job | `tileIterator` (serialize and restore) |

---

## Parallelization backends

All `tileApply` methods delegate to `future.apply::future_lapply`. The active
`future` plan controls how work is distributed.

```{r parallel, eval=FALSE}
library(future)

# Sequential (default, no parallelism)
plan(sequential)

# Local multicore (forking, fast on Linux/macOS)
plan(multicore, workers = 4L)

# Local multisession (separate R processes, works everywhere)
plan(multisession, workers = 4L)

# Mirai backend (lower overhead, faster for many workers)
plan(future.mirai::mirai_multisession, workers = 4L)

# Reset to sequential when done
plan(sequential)
```

`tileGroup` with `parallel_strategy = "groups"` parallelizes at the group
level — each `future` worker handles one group. With `parallel_strategy =
"tiles"`, each worker handles one tile within the current group.

`tileIterator` always parallelizes at the worker level: `iterSplit` divides
the tile range across `nbrOfWorkers()` sub-iterators, and each worker streams
through its assigned range independently.
