---
title: "Choosing and Creating a Tile Plan"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Choosing and Creating a Tile Plan}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(tilework)
library(terra)
```

## Overview

tilework provides four concrete tile plan classes, each suited to a different
way of thinking about how a dataset should be divided:

| Class | Tile placement | Coordinate system | Returns |
|---|---|---|---|
| `spatialTilePlan` | Uniform grid | CRS units | `SpatExtent` |
| `pixelTilePlan` | Uniform grid | Pixel indices | `integer[4]` |
| `pointTilePlan` | Arbitrary centers | CRS or pixel | `SpatExtent` or `integer[4]` |
| `freeTilePlan` | Explicit per-tile bounds | Inherent to bounds | `SpatExtent` |

The choice between them is usually straightforward:

- Working with **georeferenced rasters or vectors** where coordinates are in
  CRS units? → `spatialTilePlan`
- Working with **images** where pixel position is what matters? → `pixelTilePlan`
- Have a **list of locations** (cell centroids, tissue landmarks, sample sites)
  you want to extract fixed-size patches around? → `pointTilePlan`
- Need **variable-size tiles** driven by data density or an external
  partitioning algorithm? → `freeTilePlan` (often via `quadtreePlan()`)

---

## `spatialTilePlan`

`spatialTilePlan` divides a spatial extent into a uniform grid. Tile bounds are
returned as `SpatExtent` objects, which route directly into terra's spatial
windowing operations downstream.

Two things are required: an extent and a tile count request. The actual number
of tiles generated will be at least the requested number, arranged as squarely
as possible given the extent's aspect ratio.

```{r spat-setup}
tp_spat <- spatialTilePlan(ext = c(0, 1000, 0, 600), n = 12)

tp_spat
dim(tp_spat)   # nrow x ncol of the tile grid
length(tp_spat)
```

```{r spat-plot, fig.width=5, fig.height=3}
plot(tp_spat)
```

Padding expands each tile's bounds equally on all four sides at extraction
time. The plan's own extent is not affected.

```{r spat-pad, fig.width=5, fig.height=3}
tp_spat_padded <- tp_spat + 20
plot(tp_spat_padded, alpha = 0.4)
```

Indexing returns `SpatExtent` objects:

```{r spat-index}
tp_spat[1]       # first tile
tp_spat[2, 3]    # row 2, col 3
```

**When to use:** Any workflow where tile bounds need to be expressed in the
same coordinate system as the data — georeferenced rasters, `sf` geometries,
`SpatVector` objects.

---

## `pixelTilePlan`

`pixelTilePlan` divides an image into a uniform grid of pixel-exact tiles.
Bounds are returned as `integer[4]` vectors `c(xmin, xmax, ymin, ymax)` in
pixel index space, routing into array/matrix subsetting operations downstream.

Three things are required: the image's pixel dimensions (`pxdims`), and the
tile height and width in pixels (`nrows`, `ncols`).

```{r px-setup}
tp_px <- pixelTilePlan(pxdims = c(500, 800), nrows = 100, ncols = 100)

tp_px
dim(tp_px)
length(tp_px)
```

```{r px-plot, fig.width=5, fig.height=3}
plot(tp_px)
```

Padding for pixel plans is automatically zeroed against the top-left corner
so that padded tiles on the first row/column do not produce negative indices.

```{r px-pad}
tp_px_padded <- tp_px + 5
tp_px_padded[1]   # first tile: note xmin/ymin are still >= 1
tp_px[1]          # unpadded for comparison
```

**When to use:** Image processing workflows where there is no meaningful
spatial coordinate system — microscopy images, whole slide images, general
computer vision pipelines. Also the right choice when pixel-exact tile
boundaries matter (e.g. ML patch extraction where tile size must match model
input exactly).

---

## `pointTilePlan`

`pointTilePlan` places tiles at arbitrary (x, y) centers with a uniform tile
size. Instead of dividing a grid, you provide the locations and the plan
generates one tile per point.

This is the right choice when the locations to sample are known in advance —
cell nuclei, tissue landmarks, sampling sites, feature centroids — rather than
when you want to exhaust a spatial extent.

```{r pt-setup}
tp_pt <- pointTilePlan("spatial",
    coords = cbind(x = c(100, 250, 400, 550), y = c(80, 200, 150, 300)),
    width  = 80,
    height = 80
)

tp_pt
length(tp_pt)
```

```{r pt-plot, fig.width=5, fig.height=4}
plot(tp_pt)
```

### Coordinate modes: `input` and `output`

`pointTilePlan` has two independent coordinate mode settings:

- **`$input`** — the space in which `$coords`, `$width`, and `$height` are
  expressed. Either `"spatial"` (CRS units) or `"pixel"` (pixel indices).
- **`$output`** — the type of bounds returned by `getTile()`. Either
  `"spatial"` (returns `SpatExtent`) or `"pixel"` (returns `integer[4]`).

By default `output` matches `input`. Setting them to different values enables
cross-mode conversion.

The four combinations:

| `input` | `output` | Meaning |
|---|---|---|
| `"spatial"` | `"spatial"` | Spatial coords in, spatial bounds out |
| `"pixel"` | `"pixel"` | Pixel coords in, pixel bounds out |
| `"spatial"` | `"pixel"` | Define location in CRS, extract pixel bounds |
| `"pixel"` | `"spatial"` | Define location in pixel space, extract spatial bounds |

### Same-mode: spatial → spatial

The most common case. Coordinates, tile dims, and padding are all in CRS
units. No raster metadata needed.

```{r pt-s2s}
tp_s2s <- pointTilePlan(input = "spatial", output = "spatial",
    coords = cbind(x = c(200, 500), y = c(150, 400)),
    width  = 100,
    height = 100
)

tp_s2s[1]  # SpatExtent
```

### Same-mode: pixel → pixel

Coordinates and tile dims are in pixel indices. Useful when sampling locations
come from image analysis (e.g. detected cell positions in pixel space).

```{r pt-p2p}
tp_p2p <- pointTilePlan(input = "pixel", output = "pixel",
    coords = cbind(x = c(120L, 340L, 560L), y = c(80L, 200L, 310L)),
    width  = 51L,   # odd dims give unambiguous center pixel
    height = 51L
)

tp_p2p[2]  # integer[4]: col_min, col_max, row_min, row_max
```

Note: tile dims should be **odd** in pixel mode. With even dimensions,
`center ± width/2` is non-integer and bounds are truncated, biasing left/up.

### Cross-mode: spatial → pixel and pixel → spatial

Cross-mode conversion requires knowing how CRS coordinates map to pixel
indices, which means the raster's resolution and extent must be known.

When `getTile()` is called with a `SpatRaster`, this information is injected
automatically from the raster — you do not need to set `$rast_dims` or
`$extent` manually. They are only needed for standalone `[i]` calls without
a raster (e.g. plotting, validation).

```{r pt-cross-standalone}
f <- system.file("ex/elev.tif", package = "terra")
r <- rast(f)

# spatial input, pixel output
tp_s2p <- pointTilePlan(input = "spatial", output = "pixel",
    coords = cbind(x = c(5.8, 6.0, 6.2), y = c(49.9, 50.0, 50.1)),
    width  = 0.1,
    height = 0.1
)

# Standalone [i] requires rast_dims and extent:
tp_s2p$rast_dims <- dim(r)[1:2]
tp_s2p$extent    <- as.vector(ext(r))

tp_s2p[1]  # pixel bounds, derived from spatial coords + raster metadata
```

```{r pt-cross-getTile}
# At getTile() time, rast_dims and extent are injected from the raster
# automatically — no manual setup needed for the extraction step:
tiles <- getTile(r, tp_s2p, i = 1:3)
dim(tiles[[1]])
```

---

## `freeTilePlan` and `quadtreePlan`

`freeTilePlan` stores explicit per-tile bounds as an n × 4 matrix
(`xmin, xmax, ymin, ymax`). Unlike the grid-based plans, tile sizes can vary
freely — bounds are the canonical representation, not a formula.

The simplest use is to supply bounds directly:

```{r free-manual}
tp_free <- freeTilePlan()
tp_free$bounds <- rbind(
    c(0,  50,  0,  50),
    c(50, 100, 0,  50),
    c(0,  50,  50, 100),
    c(50, 100, 50, 100)
)

tp_free
length(tp_free)
tp_free[2]   # SpatExtent
```

```{r free-plot, fig.width=5, fig.height=4}
plot(tp_free)
```

### Adaptive decomposition with `quadtreePlan()`

The more common workflow is to let `quadtreePlan()` build the `freeTilePlan`
automatically. It starts from a coarse `tilePlan`, applies `FUN` to each tile,
and recursively splits tiles whose value exceeds `threshold`. Splitting
continues until all tiles are below the threshold, smaller than
`min_tile_size`, or `max_depth` is reached. After splitting, neighboring leaf
tiles whose combined `FUN` value stays ≤ `threshold` are merged back into a
single rectangle, reducing tile count in sparse regions.

`FUN` must return a single numeric scalar per tile — typically a count,
density, or variance. The last measured value for each leaf is stored in the
`n_records` metadata column.

```{r quadtree-setup}
# Synthetic two-cluster point cloud
set.seed(42)
pts <- terra::vect(
    cbind(x = rnorm(1000, 0, 50), y = rnorm(1000, 0, 50))
)
pts <- rbind(pts, terra::vect(
    cbind(x = rnorm(1000, 300, 50), y = rnorm(1000, 300, 50))
))

# tileApply requires on-disk data
f_pts <- tempfile(fileext = ".shp")
terra::writeVector(pts, f_pts)
pts_proxy <- terra::vect(f_pts, proxy = TRUE)

# Coarse 2×2 starting grid
tp_start <- spatialTilePlan(ext = ext(pts), n = 4L)
tp_start
```

```{r quadtree-run}
fp <- quadtreePlan(
    x             = pts_proxy,
    tiles         = tp_start,
    FUN           = function(x) length(x),   # point count per tile
    threshold     = 300L,
    min_tile_size = 10
)
fp
length(fp)
```

```{r quadtree-plot, fig.width=5, fig.height=5}
plot(fp)
```

High-density regions are subdivided into many small tiles; sparse regions
remain as large tiles (or are merged back if neighbors are also sparse).
`n_records` records the point count for each leaf:

```{r quadtree-nrecords}
summary(fp$n_records)
```

```{r quadtree-nrecords-plot, fig.width=5, fig.height=5}
plot(fp, values = "n_records")
```

The resulting `freeTilePlan` is a standard `tilePlan` and works with
`tileApply()`, `tileIterator`, and `tileGroup` like any other.

**When to use:**

- Vector/point data where density varies strongly across space.
- Any workflow where a uniform grid would waste computation in empty regions
  or exceed memory limits in dense regions.
- As the output of an external space-partitioning step (pass bounds directly
  via `$bounds<-`).

**Limitations:** `@tile_dims` is not populated, so operations that require
uniform tile dimensions (e.g. `extend = TRUE` in `getTile()`) are not
supported.

---

## Choosing tile size and padding

Regardless of plan type, two practical considerations apply:

**Tile size** controls the memory footprint per tile and, for ML workflows,
should be chosen with the model's expected input dimensions in mind. Larger
tiles mean fewer total tiles but more memory per tile. If the natural tile
size does not match the model input exactly, `terra::aggregate()` or
`terra::resample()` can resize the extracted tile inside `FUN` rather than
constraining the tile plan itself.

**Padding** handles edge effects for operations that need spatial context
(convolutions, focal statistics, buffer-based extractions). The pad value
should be at least as large as the half-width of the operation's kernel or
buffer radius. For pixel plans, padding is zero-clamped at image edges; for
spatial plans it extends beyond the plan extent, so shrink the plan extent
inward by the pad amount if you want tiles to stay within the data boundary.

```{r padding-demo, fig.width=5, fig.height=3}
tp_demo <- spatialTilePlan(ext = c(0, 100, 0, 100), n = 9)

# shrink extent inward by pad so padded tiles stay within data
pad <- 3
ext(tp_demo) <- ext(tp_demo) - pad
tp_demo <- tp_demo + pad

plot(tp_demo, alpha = 0.4)
plot(ext(c(0, 100, 0, 100)), add = TRUE, border = "red")
```

---

## Summary

```{r summary-table, echo=FALSE}
knitr::kable(data.frame(
    Plan = c("spatialTilePlan", "pixelTilePlan", "pointTilePlan", "freeTilePlan"),
    `Tile placement` = c("Uniform grid", "Uniform grid", "Arbitrary centers", "Explicit per-tile bounds"),
    `Coordinate system` = c("CRS units", "Pixel indices", "CRS or pixel (configurable)", "Inherent to bounds"),
    `Returns` = c("SpatExtent", "integer[4]", "SpatExtent or integer[4]", "SpatExtent"),
    `Primary use case` = c(
        "Georeferenced rasters and vectors",
        "Images, ML patch extraction",
        "Sampling at known locations",
        "Adaptive/variable-size tiling; quadtree decomposition"
    ),
    check.names = FALSE
))
```
